
# Interaction in linear models
```{r}
pacman::p_load(conflicted,tidyverse,car,multcomp,wrappedtools, broom)
conflicts_prefer(dplyr::select)
```
For a better understanding, data with defined effect size and interactions will be simulated and analyzed.

## No age effect, no treatment effect, interaction treatment*agegroup

```{r}
set.seed(101)
rawdata <- tibble(
  Agegroup=factor(
    rep(c('young','old'),each=40),
    levels=c('old','young')),
  Treatment=factor(
    rep(c('sham','OP'),40),
    levels = c('sham','OP'))) |>
  mutate(Outcome=rnorm(n = 80,mean = 100,sd = 10)+
           ((Treatment=='OP')*
              (Agegroup=='young'))*20)
ggplot(rawdata,aes(x=Agegroup,y=Outcome,
                   fill=Treatment))+
  geom_boxplot()+
  geom_point(position=position_dodge(width=.75))
lmout <- lm(Outcome~Agegroup*Treatment,
             data = rawdata)
tidy(lmout) |> 
  select(1:2)
anova(lmout) |>  # this is WRONG!!!
  tidy() |> slice(1:3) |> 
  mutate(p.value=formatP(p.value,ndigits=3, mark=TRUE))
  
Anova(lmout,type = 3) |> 
  tidy() |> slice(1:4) |> 
  mutate(p.value=formatP(p.value,ndigits=3, mark=TRUE))

summary(glht(model=lmout,
             linfct=mcp(Treatment='Tukey')))
summary(glht(model=lmout,
             linfct=mcp(Agegroup='Tukey')))

```
\newpage
## Age effect, no treatment effect, interaction treatment*agegroup

```{r}
set.seed(101)
rawdata <- tibble(
  Agegroup=factor(
    rep(c('young','middle','old'),each=40),
    levels=c('young','middle','old')),
  Treatment=factor(
    rep(c('sham','OP'),60),
    levels = c('sham','OP')),
  Outcome=rnorm(120,100,10)+
    (Treatment=='OP')*
    (Agegroup=='middle')*20+
    (Agegroup=='old')*20)
ggplot(rawdata,aes(x=Agegroup,y=Outcome,
                   fill=Treatment))+
  # geom_boxplot()+
  ggbeeswarm::geom_beeswarm(dodge.width = .75)+
  stat_summary(aes(color=Treatment),
               position=position_dodge(width = .75))
lmout <- lm(Outcome~Agegroup*Treatment,
             data = rawdata)
tidy(lmout) |> 
  select(1:2)
anova(lmout) |>  # this is WRONG!!!
  tidy() |> slice(1:3) |> 
  mutate(p.value=formatP(p.value,ndigits=3, mark=TRUE))
Anova(lmout,type = 3) |> 
  tidy() |> slice(1:4) |> 
  mutate(p.value=formatP(p.value,ndigits=3, mark=TRUE))
summary(glht(model=lmout,
             linfct=mcp(Treatment='Tukey')))
summary(glht(model=lmout,
             linfct=mcp(Agegroup='Tukey')))
```
\newpage
## Age effect, treatment effect, interaction treatment*agegroup

```{r}
set.seed(101)
rawdata <- tibble(
  Agegroup=factor(
    rep(c('young','middle','old'),each=40),
    levels=c('young','middle','old')),
  Treatment=factor(
    rep(c('sham','OP'),60),
    levels = c('sham','OP'))) |>
  mutate(Outcome=rnorm(120,100,10)+
           (Treatment=='OP')*
           # (Agegroup %in% c('young','middle'))*
           (Agegroup!='old')*20+
           (Agegroup=='old')*20)
ggplot(rawdata,aes(x=Agegroup,y=Outcome,
                   fill=Treatment))+
  geom_boxplot()+
  ggbeeswarm::geom_beeswarm(dodge.width = .75)
suppressWarnings(
  ggplot(rawdata,aes(x=as.numeric(Agegroup),y=Outcome,fill=Treatment))+
  ggbeeswarm::geom_beeswarm(aes(shape=Treatment),
                            alpha=.5, dodge.width = .15)+
  geom_smooth()+
  scale_x_continuous("Agegroup",breaks=1:3,
                     labels=c('young','middle','old')))

lmout <- lm(Outcome~Agegroup*Treatment,
             data = rawdata)
tidy(lmout) |> 
  select(1:2)
anova(lmout) |>  # this is WRONG!!!
  tidy() |> slice(1:3) |> 
  mutate(p.value=formatP(p.value,ndigits=3, mark=TRUE))
  
Anova(lmout,type = 3) |> 
  tidy() |> slice(1:4) |> 
  mutate(p.value=formatP(p.value,ndigits=3, mark=TRUE))
summary(glht(model=lmout,
             linfct=mcp(Treatment='Tukey')))
summary(glht(model=lmout,
             linfct=mcp(Agegroup='Tukey')))
```
Ignoring the interaction in the model is no solution. Effect sizes will be wrongly estimated:
```{r}
# falsch!!!
lmout <- lm(Outcome~Agegroup+Treatment,
             data = rawdata)
tidy(lmout) |> 
  select(1:2)
anova(lmout) |>  # this is WRONG!!!
  tidy() |> slice(1:2) |> 
  mutate(p.value=formatP(p.value,ndigits=3, mark=TRUE))
  
Anova(lmout,type = 2) |> 
  tidy() |> slice(1:2) |> 
  mutate(p.value=formatP(p.value,ndigits=3, mark=TRUE))
```
\newpage
## How to specify interaction in multivariable models

```{r}
# all possible interactions
(lm_out <- lm(mpg~(wt*gear*factor(am)*cyl),
              data=mtcars))
# only two-way interactions
(lm_out <- lm(mpg~(wt+gear+factor(am)+cyl)^2,
              data=mtcars))
#some seected interactions
(lm_out <- lm(mpg~wt*(gear+factor(am)+cyl),
              data=mtcars))
```


