# Changing structure wide \<--\> long

When working with repeated measures (e.g. follow-ups or changes over shorter periods of time) there are two typical formats:

1.  wide data:

    | ID  | Var1Time1 | Var1Time2 | Var2Time1 | Var2Time2 |
    |-----|-----------|-----------|-----------|-----------|
    | P1  |           |           |           |           |
    | P2  |           |           |           |           |
    | P3  |           |           |           |           |

2.  long data

    | ID  | Time | Var1 | Var2 |
    |-----|------|------|------|
    | P1  | 1    |      |      |
    | P1  | 2    |      |      |
    | P2  | 1    |      |      |
    | P2  | 2    |      |      |
    | P3  | 1    |      |      |
    | P3  | 2    |      |      |

While long data can be seen as the tidier version and is necessary for many statistical procedures, the wide format makes computation of differences and procedures as the t-test for dependent samples easier. Package tidyr provides the functions pivot_wider and pivot_longer for conversations between those forms. Another use-case is the plotting of several variables into ggplot facets, which can be achieved by combining those variables into a single column. Summarizing several variables and grouped data may result in a wide table with groups as rows and variables as columns, contrary to the common opposite form, another use-case.

There are many examples and explanations in the vignette:

[https://tidyr.tidyverse.org/articles/pivot.html](https://tidyr.tidyverse.org/articles/pivot.html){target="_blank"}

```{r}
pacman::p_load(conflicted,tidyverse, wrappedtools)
```

Example 1: single repeated measure

```{r}
n <- 3
wide_data <- tibble(ID = paste("P",1:n),
                    Var1 = LETTERS[1:n],
                    Var2Time1 = rnorm(n = n, mean = 100, sd = 15),
                    Var2Time2 = Var2Time1 + rnorm(n,10,5))
wide_data
long_data <- pivot_longer(
  data = wide_data,
  cols = contains("Time")
)
long_data
```

Example 2: several repeated measures

```{r}
wide_data <- tibble(ID = paste("P",1:n),
                    Var1Time1 = rnorm(n = n, mean = 100, sd = 15),
                    Var1Time2 = Var1Time1 + rnorm(n,10,5),
                    Var2Time1 = rnorm(n = n, mean = 10, sd = 2),
                    Var2Time2 = Var2Time1 + rnorm(n,0,1),
                    Var3 = LETTERS[1:n])
wide_data
#intermediate:
very_long_data <- pivot_longer(
  data = wide_data,
  cols = contains("Time"),
  names_to = c("Variable", "Time"),
  names_pattern = "(Var\\d)(Time[12])",
  values_to = "Value"
)
very_long_data
long_data <- pivot_wider(very_long_data,
                         names_from = Variable,
                         values_from = Value)
long_data
```
