# Logistic regression

Generalized linear models will be introduced using logistic regression as an example. The data set infert contains information were the outcome follows a binomial distribution.

```{r}
pacman::p_load(
  conflicted, tidyverse, wrappedtools, here, flextable,
  car, ggbeeswarm,
  pROC, rpart, rpart.plot
)
```

## Odds vs. probability

Probabilities are usually expressed xx%, 20% risk for rain means that on 20 out of 100 days like this it rains. Odds are expressed as xx:yy, 20:80 means that on 20 out of 100 days like this it rains, while on 80 days it does not. The odds are calculated as p/(1-p), where p is the probability. The log odds are calculated as log(odds).\
For many traits it is reasonable to assume a sigmoidal relationship between a risk (like LDL cholesterol) and the probability of an outcome like myocardial infarction. Those probabilities are restrained between 0 and 1. The odds are restrained at 0 but have not upper bound. The log odds are not restrained and are linearly related to the risk increase. While this may cause mental knots, it is useful mathematically and forms the basis for logistic regression, a generalized linear model for outcomes with binomial distribution.

```{r}
#| echo: false
plotdata <- tibble(
  position = 1:1000,
  `log(odds)` = seq(-10, 10,
    length.out = 1000
  ),
  odds = exp(`log(odds)`),
  p = odds / (odds + 1),
  `p [%]` = paste(p * 100, "%"),
  p_txt = paste(p * 100, 100,
    sep = " in "
  ),
  odds_txt = paste(p * 100,
    (1 - p) * 100,
    sep = " to "
  )
)
plotdata |>
  select(where(is.numeric)) |>
  pivot_longer(-position,
    names_to = "Measure"
  ) |>
  ggplot(aes(position, value)) +
  geom_point(alpha = .1) +
  geom_line(aes(group = Measure)) +
  facet_grid(
    rows = vars(Measure),
    scales = "free_y"
  )
```

## Data preparation

Before the analysis, the data set is cleaned and prepared for the analysis. Variables with 0 / 1 are coded as factor No / Yes, variable practice_problems_solved is lumped into three categories. 

```{r}
# rawdata <- infert |>
#   as_tibble() |>
#   select(-contains("stratum"))
rawdata <- 
  read_csv(here("data/python_learning_exam_performance.csv"))
head(rawdata)
```

```{r data_cleaning}
rawdata <- 
  mutate(rawdata,
         across(where(is.character), factor),
         prior_programming_experience = 
           factor(prior_programming_experience,
                  levels = c("None", "Beginner", "Intermediate", "Advanced")),
         across(where( \(x) is.numeric(x) & all(x %in% 0:1)),
                \(x) factor(x, levels = c(0,1), labels = c("No", "Yes"))),
         practice_problems_solved = 
           cut(practice_problems_solved,
               breaks = c(0, 55, 65, Inf),
               labels = c("few", "average", "many")),
  )
```

\newpage

## Exploration

```{r}
plot_data <- rawdata |> 
  count(prior_programming_experience, passed_exam) |> 
  group_by(prior_programming_experience) |> 
  mutate(
    proportion = n / sum(n),
    odds = proportion[passed_exam == "Yes"] / proportion[passed_exam == "No"],
    odds_label = paste0("Odds:\n", round(odds, 2))
  ) |> 
  ungroup()
rawdata |> 
  ggplot(aes(prior_programming_experience, fill = fct_rev(passed_exam))) +
  geom_bar(position = "fill") +
  geom_label(
    data = plot_data |> distinct(prior_programming_experience, odds_label),
    aes(x = prior_programming_experience, y = 0.02, label = odds_label),
    inherit.aes = FALSE,
    vjust = 0,
    fontface = "bold"
  ) +
  scale_y_continuous("Frequency", labels = scales::percent) +
  scale_fill_manual("Exam",
                    values = c("lightgreen", "salmon"),
                    labels = c("passed", "not passed"))
```
  
```{r}
rawdata |> 
  ggplot(aes(hours_spent_learning_per_week, 
             fill = fct_rev(passed_exam))) +
  geom_histogram(position = "fill", binwidth = 2, 
                 center = 1, closed = "left") +
  scale_y_continuous("Frequency", labels = scales::percent) +
  scale_x_continuous(breaks = seq(0,20,2)) +
  scale_fill_manual("Exam",
                    values = c("lightgreen", "salmon"),
                    labels = c("passed", "not passed"))
rawdata |> 
  ggplot(aes(age, fill = fct_rev(passed_exam))) +
  geom_histogram(position = "fill", binwidth = 5, 
                 center = 1, closed = "left") +
  scale_y_continuous("Frequency", labels = scales::percent) +
  # scale_x_continuous(breaks = seq(0,20,2)) +
  scale_fill_manual("Exam",
                    values = c("lightgreen", "salmon"),
                    labels = c("passed", "not passed"))
```

The probability of passing the exam is going up with more experience and more learning hours, as reflected in the odds of passing. Age does not seem to have a significant effect.  
  
## Build model

The model is built using `glm()` and the output is extracted and transformed. The model is tested using `Anova()` and `summary()`. The results are then prepared for plotting.

```{r}
pred_vars <- ColSeeker(namepattern = ".", 
                       exclude = c("_id", "exam", 
                                   "kaggle", "country"))
glm_formula <- reformulate(pred_vars$names, response = "passed_exam")
logreg_out <-
  glm(
    # case ~ `age [pentayears]` + education + parity_grp +
    #   induced_f + spontaneous_f,
    glm_formula, family = binomial(), data = rawdata
  )
logreg_out

# extract/transform model parameters
(ORs <- exp(logreg_out$coefficients))
(CIs <- exp(confint(logreg_out)))

# test model
(Anova_out <- Anova(logreg_out, type = 2) |>
  broom::tidy() |>
  mutate(p.value = formatP(p.value, ndigits = 5)))
## test each OR
#(sum_out <- summary(logreg_out))

sum_out <- broom::tidy(logreg_out)
```

## Create structure for ggplot

```{r}
OR_plotdata <- tibble(
  Predictor = names(ORs)[-1] |>
    # make names nicer
    str_replace_all("_", " ") |>
    str_replace_all(c(
      "([[:lower:]])([[:upper:]])" = "\\1: \\2",
      "`" = "",
      "(.)(yes)" = "\\1: \\2",
      "(solved)(.+)" = "\\1: \\2"
    )) |>
    str_to_title() |> 
    fct_inorder(),
  OR = ORs[-1],
  CI_low = CIs[-1, 1],
  CI_high = CIs[-1, 2],
  p = sum_out$p.value[-1],
  Significance = markSign(p),
  Label = paste(Predictor, Significance) |> 
    fct_inorder()
)
```

## Create forest plot

```{r}
  ggplot(OR_plotdata, aes(x = Label, y = OR)) +
  geom_pointrange(aes(ymin = CI_low, ymax = CI_high)) +
  geom_hline(yintercept = 1, linewidth = .5, linetype = 2) +
  coord_flip() +
  scale_y_continuous(limits = c(0.1, 1.5 * 10^4)) +
  xlab(NULL)

ggplot(OR_plotdata, aes(x = Predictor, y = OR)) +
  geom_pointrange(aes(ymin = CI_low, ymax = CI_high)) +
  geom_hline(yintercept = 1, linewidth = .5, linetype = 2) +
  coord_flip() +
  scale_y_log10(
    breaks = logrange_1,
    minor_breaks = logrange_123456789,
    limits = c(0.1, 2*10^4),
    labels = prettyNum) +
  geom_text(aes(label = Significance), vjust = 1.25, color = "red") +
  ggtitle("OddsRatios for passing python exam\nshown on log-scale") +
  xlab(NULL)
```

## Create predictions

```{r}
rawdata$pGLM <-
  predict(logreg_out, type = "response") # predict probability 0-1
# run ROC for cutoff
roc_out <- roc(
  response = rawdata$passed_exam,
  predictor = rawdata$pGLM
)
youden <- pROC::coords(roc_out,
  x = "best",
  best.method = "youden"
)
youden
ggroc(roc_out, legacy.axes = T) +
  geom_abline(slope = 1, intercept = 0) +
  geom_point(
    x = 1 - youden$specificity,
    y = youden$sensitivity, color = "red", size = 2
  )
# plot predictions
tempdata <-
  rawdata |>
  mutate(
    `prediction` =
      case_when(
        passed_exam == "Yes" &
          pGLM < youden$threshold ~
          "false negative",
        passed_exam == "No" &
          pGLM >= youden$threshold
        ~ "false positive",
        passed_exam == "Yes" & pGLM >= youden$threshold ~
          "correct positive",
        .default = "correct negative"
      )
  )
prediction_counts <-
  tempdata |>
  group_by(prediction) |>
  count()

rawdata |>
  mutate(
    `prediction quality` =
      case_when(
        passed_exam == "Yes" &
          pGLM < youden$threshold ~
          "false negative",
        passed_exam == "No" &
          pGLM >= youden$threshold
        ~ "false positive",
        .default = "correct"
      )
  ) |>
  ggplot(aes(passed_exam, pGLM)) +
  geom_boxplot(outlier.alpha = 0) +
  scale_y_continuous(breaks = seq(0, 1, .1)) +
  geom_beeswarm(
    alpha = .05, cex = .075,
    aes(color = `prediction quality`)
  ) +
  scale_color_manual(values = c("seagreen", "firebrick", "magenta"),
                     ) +
  guides(
    color = guide_legend(override.aes = list(alpha = .8))) +
  geom_hline(
    yintercept = c(youden$threshold, .5),
    color = "red",
    linetype = 3:4
  ) +
  annotate(
    geom = "label",
    x = 1.5, y = youden$threshold,
    label = paste(
      "Youden-cutoff:",
      roundR(youden$threshold)
    ),
    hjust = .5, vjust = 0.25
  ) +
  theme(legend.position = "bottom") +
  labs(
    caption =
      paste(prediction_counts$prediction,
        prediction_counts$n,
        collapse = "\n"
      )
  )
```

```{r}
ORreport <-
  tibble(
    Predictor = names(ORs),
    OR = ORs,
    OR_low = CIs[, 1],
    OR_high = CIs[, 2]
  ) |>
  rowwise() |>
  mutate(
    across(
      -Predictor,
      \(x) roundR(x, level = 3)
    ),
    `OR(CI)` = paste0(OR, " (", OR_low, " / ", OR_high, ")")
  ) |>
  ungroup()

ORreport |>
  slice(-1) |> 
  flextable::flextable() |>
  flextable::set_table_properties(width = 1, layout = "autofit")
cat("&nbsp;\n\n")
```

## Regression tree as alternative to glm

```{r}
#| fig.height: 6
regtree_out <- rpart(glm_formula,
  minsplit = 20, cp = .001,
  data = rawdata
)
rpart.plot(regtree_out, type = 2, tweak = 2.0, 
           varlen = 10, faclen = 5, leaf.round = 0)
```


```{r}
importance <-
  as_tibble(regtree_out$variable.importance,
    rownames = "Predictor"
  ) |>
  dplyr::rename("Importance" = 2) |>
  mutate(Predictor = fct_reorder(
    .f = Predictor,
    .x = Importance,
    .fun = min
  )) |>
  arrange(desc(Importance))
importance |>
  ggplot(aes(Predictor, Importance)) +
  geom_col() +
  coord_flip()
```

```{r}
rawdata$pRT <- predict(regtree_out)[, 2]

# pROC
roc_out_rt <- roc(
  response = rawdata$passed_exam,
  predictor = rawdata$pRT
)
youden <- pROC::coords(roc_out_rt,
  x = "best",
  best.method = "youden")

youden
ggroc(roc_out_rt, legacy.axes = T) +
  geom_abline(slope = 1, intercept = 0) +
  geom_point(
    x = 1 - youden$specificity,
    y = youden$sensitivity, color = "red", size = 2
  )

ggroc(list(RTbased = roc_out_rt, GLM_based = roc_out), legacy.axes = T) +
  geom_abline(slope = 1, intercept = 0)

ggplot(rawdata, aes(x = passed_exam, y = pRT)) +
  geom_boxplot(coef = 3) +
  scale_y_continuous(breaks = seq(from = 0, to = 1, by = .1)) +
  geom_hline(
    yintercept = c(.5, youden$threshold),
    color = c("red", "blue"), linetype = 2
  ) +
  ggbeeswarm::geom_beeswarm(alpha = .05, cex = 0.075)
```
