# k nearest neighbors knn

```{r}
pacman::p_load(conflicted,
               tidyverse,
               wrappedtools,  # just tools 
               palmerpenguins, # data
               ggforce, # for cluster plots, hulls, zoom etc
               ggbeeswarm,
               flextable,
               caret, # Classification and Regression Training Framework
               tidymodels, # alternative to caret
               gmodels, # tools for model fitting
               easystats, 
               yardstick,
               kknn)  # model performance

# conflict_scout()
conflicts_prefer(dplyr::slice, 
                 dplyr::filter,
                 palmerpenguins::penguins)

```

## Data preparation

In the penguin data set, the four numerical variables related to anatomy are `bill_length_mm`, `bill_depth_mm`, `flipper_length_mm`, and `body_mass_g`. We will use the first two variables to demonstrate the use of knn.

```{r}
rawdata <- penguins |> 
  drop_na()
predvars <- ColSeeker(rawdata,'length')
```

## Exploratory plots

Package `ggforce` provides functions to plot convex hulls, ellipses, and zoomed facets. This is helpful in exploring group separation across 2 dimensions.

```{r}
rawplot <- 
  ggplot(rawdata, 
         aes(.data[[predvars$names[1]]], 
             .data[[predvars$names[2]]], 
             color=species))+
  geom_point()  
rawplot

rawplot+
  geom_mark_hull(expand = unit(2.5,'mm'))

rawplot+
  geom_mark_ellipse(expand = unit(2.5,'mm'))
```

## Scaling

To avoid the effect of different scales / ranges of the variables, it is desirable to scale the variables. Depending on the distribution of the independent / predictor variables, there are a number of approaches to make center and spread/range of the data comparable. Here use functions from the `recipes` package from tidymodels. To compare the scaled data to the original, we'll keep the original data in the data frame.  

### Center + scale (z-score normalization)

```{r}
# Build recipe for scaling
model_formula <- 
  paste("~", paste(predvars$names, collapse = "+")) |> 
  as.formula()
rec_normalized <- recipe(model_formula, data = rawdata) |>
  step_normalize(all_predictors()) 

# Fit recipe and apply to data
penguins_normalized <- rec_normalized |>
  prep() |>
  bake(new_data = rawdata) |> 
  rename_with(~paste0(.x, "_norm"))
```

### Range transformation

```{r}
rec_ranged <- recipe(model_formula, data = rawdata) |>
  step_range(all_predictors(), min = 0, max = 1)

penguins_ranged <- rec_ranged |>
  prep() |>
  bake(new_data = rawdata) |> 
  rename_with(~paste0(.x, "_range"))
```

### Percentile transformation
```{r}
rec_quantiled <- recipe(model_formula, data = rawdata) |>
  step_percentile(all_predictors())

penguins_quantiled <- rec_quantiled |>
  prep() |>
  bake(new_data = rawdata) |> 
  rename_with(~paste0(.x, "_perc"))
```

### Visual comparison of the scaled data

```{r}
rawdata <- 
  bind_cols(rawdata, 
            penguins_normalized, 
            penguins_ranged, 
            penguins_quantiled)
rawdata |> 
  select(contains('length')) |> 
  pivot_longer(everything()) |> 
  ggplot(aes(value,fill=name))+
  geom_density()+
  facet_wrap(facets = vars(name), scales='free', nrow = 2)+
  guides(fill="none")

rawdata |> 
  ggplot(aes(bill_length_mm,bill_length_mm_norm))+ 
  geom_point() 
rawdata |> 
  ggplot(aes(bill_length_mm,bill_length_mm_perc))+ 
  geom_point()
rawdata |> 
  ggplot(aes(bill_length_mm,bill_length_mm_range))+ 
  geom_point()
```

## Modelling

### Definition of predictor variables (IV)

```{r}
predvars_std <- ColSeeker(namepattern = '_norm')
```

### Data splitting

```{r}
set.seed(2026)
data_split <- initial_split(
  rawdata,
  prop = 2/3,
  strata = species
)
traindata <- training(data_split)
testdata <- testing(data_split)
```

### Model fitting

For the training data, a model is created and used to (re-)predict the species of the test data. The `k` parameter specifies the number of neighbors to consider. Two calls to predict return classes as well as probabilities of the species.

#### Model specification and workflow

```{r}
knn_recipe <- recipe(rawdata) |>
  update_role(species, new_role = "outcome") |>
  update_role(all_of(predvars$names), new_role = "predictor")

knn_spec <- nearest_neighbor(neighbors = 5) |> 
  set_engine("kknn") |> 
  set_mode("classification")

knn_workflow <- workflow() |> 
  add_model(knn_spec)|> 
  add_recipe(knn_recipe)
```

#### Fit the model to the training data
```{r}
knn_fit <- knn_workflow |> 
  fit(data = traindata)
```
#### Predict classes

```{r}
predictions <- knn_fit |> 
  predict(new_data = testdata)
prob_preds <- knn_fit |> 
  predict(new_data = testdata, type = "prob") |> 
  rename_with(~str_remove(.x, "^.pred_"))
# Combine with original data to check accuracy
test_results <- testdata |> 
  select(species) |> 
  bind_cols(predictions,prob_preds)
```

\newpage

### Model evaluation

```{r}
CrossTable(test_results$.pred_class,
           test_results$species, prop.chisq = FALSE, prop.t = FALSE, 
           format = 'SAS')

yardstick::accuracy(data = test_results, truth=species, estimate=.pred_class)

test_results |> 
  pivot_longer(c(Adelie,Chinstrap,Gentoo), 
               values_to = 'p species', 
               names_to = 'Species_pred') |> 
  ggplot(aes(Species_pred,`p species`))+ 
  geom_violin()+ 
  geom_beeswarm(cex = .25, alpha=.25)+ 
  facet_grid(rows = vars(species))

ggplot(rawdata, aes(bill_length_mm_norm,flipper_length_mm_norm, color=sex,shape=species))+ 
  geom_point()
```

## More complex model

Now we'll try to predict the sex based on body measures and species. Species will be automatically recoded using one-hot encoding.

```{r}
model_formula <- 
  paste(" sex ~", paste(c(predvars$names, "species"), collapse = "+")) |> 
  as.formula()
penguins_preped <- recipe(model_formula, data = rawdata) |>
  step_normalize(all_numeric_predictors()) |> 
  step_dummy(all_nominal_predictors(), one_hot = TRUE) |> 
  prep() |>
  bake(new_data = rawdata) 
knn_spec <- nearest_neighbor(neighbors = 5) |> 
  set_engine("kknn") |> 
  set_mode("classification")

knn_workflow2 <- workflow() |> 
  add_model(knn_spec) |> 
  add_formula(sex ~ .)

# Fit the model to the training data
knn_fit2 <- knn_workflow2 |> 
  fit(data = penguins_preped)
# Predict classes
predictions <- knn_fit2 |> 
  predict(new_data = penguins_preped)
prob_preds <- knn_fit2 |> 
  predict(new_data = penguins_preped, type = "prob") |> 
  rename_with(~str_remove(.x, "^.pred_"))
# Combine with original data to check accuracy
all_results <- penguins_preped |> 
  select(sex) |> 
  bind_cols(predictions,prob_preds)

all_results |> 
  pivot_longer(c(female,male), 
               values_to = 'p sex', 
               names_to = 'Sex predicted') |> 
  ggplot(aes(`Sex predicted`,`p sex`))+ 
  geom_violin()+ 
  geom_beeswarm(cex = .25, alpha=.25)+ 
  facet_grid(rows = vars(sex))

yardstick::accuracy(data = all_results, truth=sex, 
                    estimate=.pred_class) 
yardstick::sensitivity(data = all_results, truth=sex, 
                       estimate=.pred_class, 
                       event_level="second") 
yardstick::specificity(data = all_results, truth=sex, 
                       estimate=.pred_class, 
                       event_level="second") 
yardstick::ppv(data = all_results, truth=sex, 
               estimate=.pred_class, 
               event_level="second") 
CrossTable(all_results$.pred_class,
           all_results$sex, prop.chisq = F, prop.t = F, 
           format = 'SPSS')

caret::confusionMatrix(all_results$.pred_class, all_results$sex,
                positive = "male")
```
