# Regression and classification trees

```{r}
#| output: false
#| message: false
pacman::p_load(conflicted,
               tidyverse,
               wrappedtools, 
               palmerpenguins,
               rpart, rpart.plot,
               randomForest,
               caret, gmodels,
               tidymodels,
               future.mirai,
               flextable,
               ggbeeswarm)

# conflict_scout()
conflicts_prefer(dplyr::slice, 
                 dplyr::filter,
                 palmerpenguins::penguins)
rawdata <- penguins |> 
  na.omit()
rawdata <- mutate(rawdata,
                  ID=paste('P', 1:nrow(rawdata))) |> 
  select(ID, everything())
```

### Graphical exploration

When looking at the 2 measures `bill_length_mm` and `flipper_length_mm`, we can see that the 3 species of penguins are separated to a large extend, with the Gentoos having longer flippers than Adelies and Chinstraps, and Adelies and Chinstraps separated by smaller / longer bills. We can define a rough classification rule based on that just by eyeballing the data.

```{r}
predvars <- ColSeeker(namepattern = 'length')
rawplot <- 
  ggplot(rawdata, 
         aes(.data[[predvars$names[1]]], 
             .data[[predvars$names[2]]], color=species))+
  geom_point()  
rawplot+
  geom_hline(yintercept = 206)+
  geom_vline(xintercept = 44)
```

### Modelling

A regression tree is following the same logic as the classification rule we defined above, but in a more systematic way. It is a recursive partitioning algorithm that splits the data into subsets that are as homogeneous as possible with respect to the target variable. The algorithm is greedy, meaning that it makes the best split at each step, without considering the impact of the split on future steps. This can lead to over-fitting, so we need to be careful with the depth of the tree. As each split is testing each variable separately, there is no need for scaling. Accordingly, the results are easily comprehensible as they relate to the actual measurements. Beside the classification rules, the output contains information on variable importance as well.

#### Direct call to rpart

```{r}
rpart_formula <- paste('species',
                       paste(predvars$names,
                             collapse='+'),
                       sep='~') |> 
  as.formula()
rpart_out <- rpart(formula = rpart_formula,
                   data = rawdata)
prp(rpart_out,
    type = 4,
    extra = 104,
    digits=4,
    fallen.leaves = TRUE)
importance <- tibble(
  Variable=names(rpart_out$variable.importance),
  Importancescore=rpart_out$variable.importance)
ggplot(importance, aes(x=Variable,y=Importancescore))+
  geom_col()+
  coord_flip()
```

#### Tidy-models approach

To make model more interesting, we add 2 more variables. We'll switch from the direct call to rpart to the tidy-models approach, as this will allow tuning and better validation strategies. Data is split into training and test sets using cross-validation. As a 1st step predictors are defined:

```{r}
predvars <- ColSeeker(namepattern = c('_'))
```

Validation is defined as 5 times repeated 5-fold cross-validation, stratified by the target variable to ensure that each fold has a similar distribution of classes. Repeating the cross-validation process multiple times helps to get a more reliable estimate of the model's performance and reduces the variance of the performance estimates.

```{r}
tree_folds <- vfold_cv(rawdata, v = 5, strata = species,repeats = 5 )
```

Data preparation includes one-hot encoding of the categorical predictor.

```{r}
tree_recipe <- recipe(rawdata) |>
  update_role(species, new_role = "outcome") |>
  update_role(all_of(predvars$names), new_role = "predictor") |> 
  step_dummy(all_nominal_predictors(), one_hot = TRUE)
```

Tuning parameters are hard-coded in a grid of parameters to search over. The `expand.grid()` function is used to create a grid of all combinations of the specified parameters.

```{r}
tree_grid <- expand.grid(tree_depth = seq(2, 6, by = 2),
                         cost_complexity = c(0.01, 0.001),
                         min_n = seq(5, 20, by = 5)) |>
  as_tibble()
```

The `tune()` function is used to indicate which parameters we want to tune.

```{r}
tree_spec <- decision_tree(cost_complexity = tune(),
                           tree_depth = tune(),
                           min_n = tune()) |>
  set_engine("rpart", model = TRUE) |>
  set_mode("classification")
tree_workflow <- workflow() |> 
  add_recipe(tree_recipe) |>
  add_model(tree_spec)
```

For the model fitting, parallel computing is used to speed up the tuning process, as it can be computationally intensive. The `plan(mirai_multisession)` function from the `future.mirai` package is used to set up a parallel processing plan that utilizes multiple CPU cores. The `availableCores()` function is used to determine the number of available cores, and we subtract 2 from that number to leave some resources free for other tasks. After the tuning process is complete, we switch back to sequential processing with `plan(sequential)`.

```{r}
# Fit the model to the data
plan(mirai_multisession, workers = availableCores()-2) # Use all but 2 cores for parallel processing
tree_results <- tree_workflow |> 
  tune_grid(
    resamples = tree_folds,
    grid = tree_grid,
    metrics = metric_set(roc_auc, accuracy)
  )
plan(sequential)
```

See the top performing models

```{r}
tree_results |> 
  show_best(metric = "accuracy") |> 
  select(-.config, -.estimator) |> 
  mutate(across(where(is.numeric), ~roundR(.x,5,drop0 = TRUE))) |> 
  flextable()

# Select the best parameters based on your preferred metric
tree_best <- tree_results |> 
  select_best(metric = "accuracy")
```

Visualize the tuning results

```{r}
tree_results |>
  collect_metrics() |>
  ggplot(aes(x = tree_depth, y = mean, color = as.factor(cost_complexity))) +
  geom_line(position = position_dodge(.25)) +
  geom_point(position = position_dodge(.25)) +
  scale_color_discrete("cost complexity") +
  facet_grid(cols = vars(.metric), rows = vars(min_n),
             labeller = "label_both") +
  theme_bw()
tree_results |>
  collect_metrics(summarize = FALSE) |>
  filter(.metric == "accuracy") |>
  ggplot(aes(x = factor(tree_depth), y = .estimate)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_beeswarm(alpha = 0.3,
                aes(color = .config == tree_best$.config)) +
  facet_grid(rows = vars(min_n), cols = vars(cost_complexity), #scales = "free",
             labeller = "label_both") +
  scale_color_manual(
    "best model",
    values = c("darkgrey", "orange"),
    labels = c("no", "yes")) +
  xlab("tree depth") +
  theme_bw()
```

Update the workflow with those best parameters

```{r}

final_workflow <- tree_workflow |> 
  finalize_workflow(tree_best)

# Fit to the entire dataset (or use test split)
final_fit <- final_workflow |> 
  fit(data = rawdata)

test_predicted <- final_fit |> 
  augment(new_data = rawdata) |>
  rename_with(~str_remove(.x, "^.pred_"), starts_with(".pred_"))
```

### Model evaluation

class-based metrics:

| Metric | Function | Meaning |
|----|----|----|
| Accuracy | accuracy() | The percentage of total predictions that were correct. Simple, but can be misleading if classes are imbalanced |
| Kappa | kap() | Similar to accuracy, but accounts for the possibility of the agreement occurring by chance |
| Sensitivity (Recall) | sens() | Out of all actual positives, how many did the model find? (True Positive Rate) |
| Specificity | spec() | Out of all actual negatives, how many did the model correctly identify? (True Negative Rate) |
| Precision (PPV) | precision() | Out of all rows the model claimed were positive, how many actually were? |
| F-Measure | f_meas() | The harmonic mean of Precision and Recall. Great for finding a balance between the two |

probability-based metrics:

| Metric | Function | Meaning |
|----|----|----|
| ROC AUC | roc_auc() | Measures the model's ability to distinguish between classes across all possible thresholds |
| PR AUC | pr_auc() | Area under the Precision-Recall curve. Preferred over ROC AUC when the "Positive" class is very rare |
| Log Loss | mn_log_loss() | Measures how "close" the predicted probability is to the actual binary label (0 or 1). Heavily penalizes being "confidently wrong" |
| Brier Score | brier_class() | The mean squared error of the probability predictions. A lower score is better |

```{r}
CrossTable(test_predicted$class,
           test_predicted$species,
           prop.chisq = F, prop.t = F,
           format = 'SPSS')

confusionMatrix(test_predicted$class,
                test_predicted$species)

tidy_metrics <- metric_set(accuracy, kap, f_meas,roc_auc, mn_log_loss)
tidy_metrics(test_predicted, truth = species, estimate = class,
             Adelie, Chinstrap, Gentoo) |> 
  flextable()
```
