# Random forrests

```{r}
#| output: false
#| message: false
pacman::p_load(
  conflicted,
  tidyverse,
  ggbeeswarm,
  wrappedtools, # just tools
  palmerpenguins, # data
  caret, # Classification and Regression Training
  gmodels, # tools for model fitting
  easystats,
  yardstick, # model performance
  xgboost,
  tidymodels,
  vip,
  flextable,
  future.mirai
) # XGBoost
conflicts_prefer(
  dplyr::slice,
  dplyr::filter,
  dplyr::lag,
  palmerpenguins::penguins
)

rawdata <- penguins |>
  na.omit() |>
  mutate(year = factor(year))
```

Random forrests are an ensemble method that builds multiple decision trees and averages their predictions. This reduces the risk of overfitting and increases the accuracy of the model. The random part comes from the fact that each tree is built on a random subset of the data and a random subset of the variables. The number of trees and the number of variables to consider at each split are hyper-parameters that need to be tuned. (more on tuning in chapter caret)

## Modelling

```{r}
rf_folds <- vfold_cv(rawdata, v = 5, strata = species, repeats = 5)
predvars <- ColSeeker(namepattern = c("_"))
rf_recipe <- recipe(rawdata) |>
  update_role(species, new_role = "outcome") |>
  update_role(all_of(predvars$names), new_role = "predictor")

rf_grid <- expand.grid(mtry = 2:4) |>
  as_tibble()
rf_spec <- rand_forest(
  trees = 1000,
  mtry = tune()
) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")
rf_workflow <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_spec)

# Fit the model to the data
plan(mirai_multisession, workers = availableCores() - 2) # Use all but 2 cores for parallel processing
rf_results <- rf_workflow |>
  tune_grid(
    resamples = rf_folds,
    grid = rf_grid,
    metrics = metric_set(roc_auc, accuracy)
  )
plan(sequential)

# See the top performing models
rf_results |>
  show_best(metric = "accuracy") |>
  select(-.config, -.estimator) |>
  mutate(across(where(is.numeric), ~ roundR(.x, 5, drop0 = TRUE))) |>
  flextable()

# Select the best parameters based on your preferred metric
rf_best <- rf_results |>
  select_best(metric = "accuracy")

# Visualize the tuning results
rf_results |>
  collect_metrics() |>
  ggplot(aes(x = mtry, y = mean)) +
  geom_line() +
  geom_point() +
  facet_grid(rows = vars(.metric)) +
  theme_bw()

rf_results |>
  collect_metrics(summarize = FALSE) |>
  filter(.metric == "accuracy") |>
  ggplot(aes(x = factor(mtry), y = .estimate)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_beeswarm(
    alpha = 0.3,
    aes(color = .config == rf_best$.config)
  ) +
  scale_color_manual(
    "best model",
    values = c("darkgrey", "orange"),
    labels = c("no", "yes")
  ) +
  theme_bw()

# Update the workflow with those best parameters
final_workflow <- rf_workflow |>
  finalize_workflow(rf_best)

# Fit to the entire dataset (or use test split)
final_fit <- final_workflow |>
  fit(data = rawdata)

test_predicted <- final_fit |>
  augment(new_data = rawdata) |>
  rename_with(~ str_remove(.x, "^.pred_"), starts_with(".pred_"))
```

## Model evaluation

```{r}
tidy_metrics <- metric_set(accuracy, kap, f_meas, roc_auc, mn_log_loss)
tidy_metrics(test_predicted,
  truth = species, estimate = class,
  Adelie, Chinstrap, Gentoo
) |>
  flextable()
```
