# Boosted regression trees

```{r}
pacman::p_load(conflicted,
               tidyverse,
               ggbeeswarm,
               wrappedtools,  # just tools 
               palmerpenguins, # data
               caret, # Classification and Regression Training
               gmodels, # tools for model fitting
               easystats, 
               yardstick,# model performance
               xgboost,
               tidymodels)  # XGBoost
conflicts_prefer(dplyr::slice, 
                 dplyr::filter,
                 dplyr::lag,
                 palmerpenguins::penguins)

rawdata <- penguins |> 
  na.omit() |> 
  mutate(year = factor(year)
         )
```

## Preprocessing
one_hot encoding transforms factors in dummy columns for EVERY level, recommended for tree-based models as they can handle the resulting collinearity, and it is more intuitive to interpret the importance of each level. The `recipe` package provides a convenient way to do this, but it can also be done manually with `dummyVars()` from the `caret` package. The resulting columns are added to the original data frame, and the original factor columns can be removed if desired.
For linear models, one-hot encoding is not recommended as it can lead to multicollinearity issues.
```{r}
# one-hot-encoding sex + island 
# tmp <- dummyVars(~sex+island, rawdata,fullRank = F)
#predict(tmp,rawdata) |> view()
# rawdata <- bind_cols(rawdata,
#                      predict(tmp,rawdata))

rawdata <- 
  recipe(species ~ ., data = rawdata) |>
  step_dummy(all_nominal_predictors(),one_hot = TRUE) |> 
  prep() |>
  bake(new_data = rawdata)
cn()
predvars <- ColSeeker(rawdata,
                      c('_.+_','sex.+',
                        "island.+","year_"))
predvars$names

boostdata <- rawdata |>
  select(predvars$names) |>
  as.matrix()
```

```{r}
xgb_out <- xgboost(
  x = boostdata,
  y = rawdata$species,
  nrounds = 100,
  objective = "multi:softprob") # requests class probabilities
prediction <- predict(xgb_out,boostdata) |> 
  as_tibble() |> 
  mutate(predicted= # find highest prob
           case_when(Adelie>=Chinstrap &
                       Adelie>=Gentoo ~ 'Adelie',
                     Gentoo>=Chinstrap ~ 'Gentoo',
                     !is.na(Chinstrap) ~ 'Chinstrap'
           ) |> 
           factor())

slice_sample(prediction, n=20)
```

```{r}
train_res <- bind_cols(rawdata,prediction)
train_res |> 
  pivot_longer(c(Adelie,Chinstrap,Gentoo),
               values_to = 'p species',
               names_to = 'Species prediction') |> 
  ggplot(aes(`Species prediction`,`p species`))+
  # geom_violin()+
  geom_beeswarm(cex = .25, alpha=.25)+
  facet_grid(rows = vars(species),
             labeller='label_both')

CrossTable(train_res$predicted,
           train_res$species,
           prop.chisq = F, prop.t = F,
           format = 'SPSS')

yardstick::accuracy(data = train_res,
                    truth=species,
                    estimate=predicted)
yardstick::specificity(data = train_res,
                    truth=species,
                    estimate=predicted)

confusionMatrix(train_res$predicted,
                train_res$species)  

xgb.importance(model = xgb_out) |> 
  as_tibble() |> 
  ggplot(aes(reorder(Feature, Gain), Gain))+
  geom_col()+
  coord_flip()+
  labs(x='Feature', y='Gain')
```

```{r}
# splitting ####
set.seed(2026)
data_split <- initial_split(
  rawdata,
  prop = 2/3,
  strata = species
)
traindata <- training(data_split)
testdata <- testing(data_split)
# splitted matrix-objects
trainobject <- 
  xgb.DMatrix(
    data=traindata |> 
      select( -contains("species")) |> 
      data.matrix(),
    label=traindata$species |> 
      as.numeric() -1)
testobject <- 
  xgb.DMatrix(
    data=testdata |>
      select(-contains("species")) |> 
      data.matrix(),
    label=testdata$species |> 
      as.numeric() -1)
```

```{r}
watchlist = list(train=trainobject, 
                 test=testobject)
# preliminary fitting to find nrounds
xgb.Train0 <- xgb.train(
  data = trainobject, 
  params = list(max.depth = 5,
                num_class=3,
                objective = "multi:softprob"),  # requests class probabilities
  evals = watchlist, #objects used in eval
  nrounds = 10^2) # set high, best will be checked afterwards
bestround <- 
  attr(xgb.Train0, "evaluation_log") |> 
  as_tibble() |> 
  filter(test_mlogloss==min(test_mlogloss)) |> 
  pull(iter)
bestround
# v2 with threshold for improvement
bestround <- 
  attr(xgb.Train0, "evaluation_log") |> 
  as_tibble() |> 
  filter(test_mlogloss-lead(test_mlogloss)>.001) |> 
  pull(iter)|> 
  max()
bestround
```

```{r}
# modelling
xgb.Train <- xgb.train(
  data = trainobject, 
  params = list(objective = "multi:softprob", 
                max.depth = 5,
                num_class=3),
  nrounds = bestround)
prediction <- predict(xgb.Train,testobject) |> 
  as_tibble() |>
  rename("Adelie"=V1,"Chinstrap"=V2,"Gentoo"=V3) |>
  mutate(predicted= # find highest prob
           case_when(Adelie>=Chinstrap &
                       Adelie>=Gentoo ~ 'Adelie',
                     Gentoo>=Chinstrap ~ 'Gentoo',
                     !is.na(Chinstrap) ~ 'Chinstrap'
           ) |> 
           factor())
train_res <- bind_cols(testdata,prediction)
```

```{r}
train_res |> 
  pivot_longer(c(Adelie,Chinstrap,Gentoo),
               values_to = 'p species',
               names_to = 'Species prediction') |> 
  ggplot(aes(`Species prediction`,`p species`))+
  geom_violin()+
  geom_beeswarm(cex = .5, alpha=.25)+
  facet_grid(rows = vars(species))

CrossTable(train_res$predicted,
           train_res$species,
           prop.chisq = F, prop.t = F,
           format = 'SPSS')

yardstick::accuracy(data = train_res,
                    truth=species,
                    estimate=predicted)

confusionMatrix(train_res$predicted,
                train_res$species)  

importance <- 
  xgb.importance(model = xgb.Train) |> 
  as_tibble() |> 
  arrange(Gain) |> 
  mutate(Feature=fct_inorder(Feature))

importance |> 
  ggplot(aes(Feature,Gain))+
  geom_col(aes(fill=Gain))+
  coord_flip()+
  guides(fill="none")

plotfeatures <- 
  slice_max(importance,Gain,n=2) |> 
  pull(Feature) |> 
  as.character()
# tail(importance$Feature,2) |> 
#  rev() |> 
#  as.character()

plotfeatures2 <- 
  xgb.importance(model = xgb.Train)[[1]][1:2] 

impo <- xgb.importance(model = xgb.Train)
plotfeatures_first <- 
  impo$Feature[1] 
plotfeatures_2nd <- 
  impo$Feature[2] 

ggplot(rawdata,aes(.data[[plotfeatures[1]]],
                   .data[[plotfeatures[2]]],
                   color=species, shape=species))+
  geom_point()+
  xlab(paste("most important:",
             plotfeatures[1],sep = "\n"))+
  ylab(paste("2nd most important:",
             plotfeatures[2],sep = "\n"))

# importance per species
xgb.importance(model = xgb.Train, 
               trees = seq(from=1, by=3, 
                           length.out=bestround))
xgb.importance(model = xgb.Train, 
               trees = seq(from=2, by=3, 
                           length.out=bestround))
xgb.importance(model = xgb.Train, 
               trees = seq(from=3, by=3, 
                           length.out=bestround))
```
