# Boosted regression trees

```{r}
pacman::p_load(conflicted,
               tidyverse,
               ggbeeswarm,
               wrappedtools,  # just tools 
               palmerpenguins, # data
               caret, # Classification and Regression Training
               gmodels, # tools for model fitting
               easystats, 
               yardstick,# model performance
               xgboost,
               tidymodels,
               vip,
               flextable,
               future.mirai)  # XGBoost
conflicts_prefer(dplyr::slice, 
                 dplyr::filter,
                 dplyr::lag,
                 palmerpenguins::penguins)

rawdata <- penguins |> 
  na.omit() |> 
  mutate(year = factor(year)
         )
```

## Direct modelling
### Preprocessing
one_hot encoding transforms factors in dummy columns for EVERY level, recommended for tree-based models as they can handle the resulting collinearity, and it is more intuitive to interpret the importance of each level. The `recipe` package provides a convenient way to do this. The resulting columns are added to the original data frame, and the original factor columns can be removed if desired.
For linear models, one-hot encoding is not recommended as it can lead to multicollinearity issues.
```{r}
rawdata <- 
  recipe(species ~ ., data = rawdata) |>
  step_dummy(all_nominal_predictors(),one_hot = TRUE) |> 
  prep() |>
  bake(new_data = rawdata)
cn()
predvars <- ColSeeker(rawdata,
                      c("_.+_","sex.+",
                        "island.+","year_"))
predvars$names

boostdata <- rawdata |>
  select(predvars$names) |>
  as.matrix()
```
### Modelling
```{r}
xgb_out <- xgboost(
  x = boostdata,
  y = rawdata$species,
  nrounds = 100,
  objective = "multi:softprob") # requests class probabilities
prediction <- predict(xgb_out,boostdata) |> 
  as_tibble() |> 
  mutate(predicted= # find highest prob
           case_when(Adelie>=Chinstrap &
                       Adelie>=Gentoo ~ "Adelie",
                     Gentoo>=Chinstrap ~ "Gentoo",
                     !is.na(Chinstrap) ~ "Chinstrap"
           ) |> 
           factor())

slice_sample(prediction, n=10)
```

## Tidy model approach


```{r}
predvars <- ColSeeker(namepattern = c("_mm","_g", "island", "sex"))

xgb_folds <- vfold_cv(rawdata, v = 5, strata = species,repeats = 5 )

xgb_recipe <- recipe(rawdata) |>
  update_role(species, new_role = "outcome") |>
  update_role(all_of(predvars$names), new_role = "predictor")

xgb_grid <- grid_space_filling(
  tree_depth(range = c(5, 10)), # Searches between 2 and 10
  trees(range = c(100, 200)), 
  learn_rate(),
  loss_reduction(),
  min_n(range = c(10,20)),
  finalize(mtry(), rawdata), 
  size = 20
)
xgb_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  min_n = tune(),
  loss_reduction = tune(),
  mtry = tune(),
  learn_rate = tune()
) |>
  set_engine("xgboost") |>
  set_mode("classification")
xgb_workflow <- workflow() |> 
  add_recipe(xgb_recipe) |>
  add_model(xgb_spec)

# Fit the model to the data
plan(mirai_multisession, workers = availableCores()-2) # Use all but 2 cores for parallel processing
xgb_results <- xgb_workflow |> 
  tune_grid(
    resamples = xgb_folds,
    grid = xgb_grid,
    metrics = metric_set(roc_auc, accuracy)
  )
plan(sequential)

# See the top performing models
xgb_results |> 
  show_best(metric = "accuracy") |> 
  select(-.metric, -.estimator) |> 
  flextable()


# Select the best parameters based on your preferred metric
xgb_best <- xgb_results |> 
  select_best(metric = "accuracy")

# Visualize the tuning results
xgb_results |> 
  collect_metrics() |>
  filter(.metric == "accuracy",
         mean>=median(mean)) |>
  arrange(desc(mean)) |>  
  ggplot(aes(x = trees, y = mean, 
             color = min_n, shape = factor(mtry))) +
#  geom_line(position = position_dodge(.25)) +
  geom_point(position = position_dodge(.25)) +
  theme_bw()

xgb_results |> 
  collect_metrics(summarize = FALSE) |>
  filter(.metric == "accuracy",
         mtry == xgb_best$mtry,
         trees == xgb_best$trees,
         min_n == xgb_best$min_n,
         tree_depth == xgb_best$tree_depth) |>
  ggplot(aes(x = factor(tree_depth), y = .estimate)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_beeswarm(alpha = 0.5, size =2) +
  xlab("tree depth") +
  theme_bw()

# Update the workflow with those best parameters
final_workflow <- xgb_workflow |> 
  finalize_workflow(xgb_best)

# Fit to the entire dataset (or use test split)
final_fit <- final_workflow |> 
  fit(data = rawdata)

test_predicted <- final_fit |> 
  augment(new_data = rawdata) |>
  rename_with(~str_remove(.x, "^.pred_"), starts_with(".pred_"))
```
 
### Model evaluation


```{r}
yardstick::conf_mat(test_predicted, truth = species, estimate = class)

tidy_metrics <- metric_set(accuracy, kap, f_meas,roc_auc, mn_log_loss)
tidy_metrics(test_predicted, truth = species, estimate = class,
             Adelie, Chinstrap, Gentoo) |> 
  flextable()

```

```{r}
test_predicted |> 
  pivot_longer(c(Adelie,Chinstrap,Gentoo), 
               values_to = 'p species', 
               names_to = 'Species_pred') |> 
  ggplot(aes(Species_pred,`p species`))+ 
  geom_violin()+ 
  geom_beeswarm(cex = .35, alpha=.25)+ 
  facet_grid(rows = vars(species))

```


```{r}
final_fit |> 
  extract_fit_engine() |> 
  vip()

final_fit |> 
  extract_fit_engine() |> 
  xgb.importance() |> 
  as_tibble() |> 
  ggplot(aes(reorder(Feature, Gain), Gain))+
  geom_col()+
  coord_flip()+
  labs(x="Feature", y="Gain")
```
