# Linear discriminant analysis LDA

LDAs are used to find a linear combination of features that characterizes or separates two or more classes of objects or events. The resulting combination may be used as a linear classifier, or, more commonly, for dimensionality reduction before later classification. LDA is comparable to PCA, but instead of finding the component axes that maximize the variance of our data (PCA), we are interested in the axes that maximize the separation between multiple classes (LDA).

```{r}
#| output: false
#| message: false
pacman::p_load(
  conflicted,
  tidyverse,
  wrappedtools,
  here,
  palmerpenguins,
  ggfortify, GGally,
  MASS,
  caret,
  tidymodels, MASS, discrim
)

# conflict_scout()
conflicts_prefer(
  dplyr::select,
  dplyr::filter,
  palmerpenguins::penguins
)
rawdata <- penguins |>
  na.omit()
```


```{r}
predvars <- ColSeeker(namepattern = c("_mm", "_g"))
norm_rec <-
  recipe(rawdata) |>
  update_role(species, new_role = "outcome") |>
  update_role(all_of(predvars$names), new_role = "predictor") |>
  step_normalize(all_numeric_predictors())
lda_data <-
  norm_rec |>
  prep() |>
  bake(new_data = rawdata)

lda_formula <- paste("species",
  paste(predvars$names, collapse = "+"),
  sep = "~"
) |>
  as.formula()
```

Direct call from MASS package

```{r}
lda_out <- lda(lda_formula, data = lda_data)
lda_out$prior
lda_out$svd^2 / sum(lda_out$svd^2) # explained variance
lda_out$scaling
lda_pred <- predict(lda_out)
lda_plotdata <-
  lda_pred$x |>
  as_tibble() |>
  cbind(rawdata |> select(species))
lda_plotdata |>
  ggplot(aes(LD1, LD2, color = species)) +
  geom_point()

confusionMatrix(lda_pred$class, rawdata$species)
```

tidymodels approach

```{r}
lda_spec <- discrim_linear() |>
  set_mode("classification") |>
  set_engine("MASS")
lda_workflow <- workflow() |>
  add_recipe(norm_rec) |>
  add_model(lda_spec)

lda_fit <-
  lda_workflow |>
  fit(data = rawdata)

lda_coefficients <- lda_fit |>
  extract_fit_engine() |>
  # For MASS engine, coefficients are in the 'scaling' element
  getElement("scaling")

print(lda_coefficients)

lda_data <-
  bind_cols(
    lda_data,
    lda_fit |>
      extract_fit_engine() |>
      predict(newdata = lda_data) |>
      pluck("x") |>
      as_tibble()
  )

ggplot(lda_data, aes(x = LD1, y = LD2, color = species)) +
  geom_point(alpha = 0.7, size = 2)
```


```{r}
# data(bordeaux)
bordeaux <- readxl::read_excel(here("Data/bordeaux.xlsx")) |>
  mutate(quality = factor(quality,
    levels = c("bad", "medium", "good")
  )) |>
  select(-year)
lda_rec <-
  recipe(bordeaux) |>
  update_role(everything(), new_role = "predictor") |>
  update_role(quality, new_role = "outcome") |>
  step_normalize(all_numeric_predictors())
lda_spec <- discrim_linear() |>
  set_mode("classification") |>
  set_engine("MASS")
lda_workflow <- workflow() |>
  add_recipe(lda_rec) |>
  add_model(lda_spec)

lda_fit <-
  lda_workflow |>
  fit(data = bordeaux)

lda_coefficients <- lda_fit |>
  extract_fit_engine() |>
  # For MASS engine, coefficients are in the 'scaling' element
  getElement("scaling")

print(lda_coefficients)

bordeaux <-
  bind_cols(
    bordeaux,
    lda_fit |>
      extract_fit_engine() |>
      predict(newdata = bordeaux) |>
      pluck("x") |>
      as_tibble()
  )

ggplot(bordeaux, aes(x = LD1, y = LD2, color = quality)) +
  geom_point(alpha = 0.7, size = 2)
```
